En garde! Unguarded iteration for reversible computation in the Delay monad\thanks{Supported by organization x.}

### 2019-04-26 mpc19/paper.tex
Added missing } in bibtex library.
Updated title, minor edits.

\begin{document}
%
\title{Type-theoretical Semantics of \Pio\ via the Delay Monad \thanks{Supported by organization x.}}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Robin Kaarsgaard \inst{1}\orcidID{0000-0002-7672-799X} \and
Niccol\`o Veltri\inst{2}\orcidID{0000-0002-7230-3436}}
%
\authorrunning{Kaarsgaard and Veltri}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{DIKU, Department of Computer Science, University of
    Copenhagen\\ \email{robin@di.ku.dk} \and
Department of Computer Science, IT University of
    Copenhagen\\ \email{nive@itu.dk}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
Reversible computation studies computations which exhibit both forward and
backward determinism. Among others, it has been studied for half a century for
its applications in low-power computing, and forms the basis for quantum
computing.

@@

In this paper, we extend the two-level calculus of finitary \Pio{} to one for
full \Pio{} (i.e., with both recursive types and iteration by means of a
\trace{} combinator) using the \Delay{} monad, which can be regarded as a
``computability-aware'' analogue of the usual $\mathsf{Maybe}$ monad for
partiality. This yields a calculus of iterative (and possibly nonterminating)
reversible programs acting on user-defined dynamic data structures together
with a calculus of certified program equivalences between these programs.

\todo{Niccol\`o, anything to add? Feel free to edit.}

@@

\section{TODO} % (fold)
\label{sec:todo}
Robin:
\begin{itemize}
  \item Abstract.
  \item Introduction.
  \item Make pictures! 
  \begin{itemize}
    \item Hasegawa trace from iterator.
  \end{itemize}
  \item Conclusion
  \begin{itemize}
    \item Related work: Kleisli category of the Delay monad is a join
    restriction category, inverse category of a join restriction category is a
    join inverse category; already known to be dagger-traced under certain
    conditions.
    \item Future work: Semantics of time invertible programming. Extension of
    the work on the Kleisli category of the Delay monad beyond $\Set$ to other
    (more general) categories, e.g. topoi.
  \end{itemize}
\end{itemize}
Niccolò: 
\begin{itemize}
  \item Fix the syntax to be closer to ordinary presentations (but do describe 
  how it is realized in Agda).
  \item Finish the section about trace and iteration.
  \item Remind the reader about double lines of coinductive definitions.
  \item Swap types in Elgot iterator.
\end{itemize}
Both:
\begin{itemize}
  \item \textbf{Important:} Come up with a punnier title! (Something about 
  delays and daggers).
\end{itemize}
% section todo (end)

\section{Introduction}\label{sec:intro}
Reversible computation is an emerging computation paradigm encompassing
computations that are not just deterministic when executed the \emph{forward}
direction, but also in the \emph{backward} direction. While this may seem
initially obscure, reversible computation forms the basis for quantum
computing, and has seen applications in a number of different areas such as
low-power computing~\cite{Landauer61}, robotics~\cite{SchultzLE18}, discrete
event simulation~\cite{Schordan15}, and the simultaneous construction of
parser/pretty printer pairs~\cite{RendelO10}. Like classical computing, it has
its own automata~\cite{Bennett73}, circuit model~\cite{DeVos10}, machine
architectures, programming
languages~\cite{JamesS12,JamesS14,YokoyamaG07,Schultz18,JacobsenKT18}, semantic
metalanguages~\cite{Giles,KaarsgaardAG17,KaarsgaardG18}, and so on.

@@

Throughout the paper, we reason constructively in Martin-L\"of type
theory. Classically, the delay monad is isomorphic to the maybe monad
$\Maybe\,X \equiv X + 1$,
and thus just a complication of something that can be expressed much
simpler. But constructively they are very different. In particular,
it is impossible to define a well-behaved trace combinator for the
maybe monad without assuming classical principles such as the law of
excluded middle.

We have fully formalized the development of the paper in the
dependently typed programming language Agda~\cite{Agda}. The code is
available online at \\ \url{https://github.com/niccoloveltri/pi0-agda}. The
formalization uses Agda 2.6.0.

>>>>

\begin{document}
%
\title{En garde! Unguarded iteration for reversible computation in the Delay monad\thanks{Supported by organization x.}}
%
\titlerunning{Unguarded iteration for reversible computation in the Delay monad}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Robin Kaarsgaard \inst{1}\orcidID{0000-0002-7672-799X} \and
Niccol\`o Veltri\inst{2}\orcidID{0000-0002-7230-3436}}
%
\authorrunning{Kaarsgaard and Veltri}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{DIKU, Department of Computer Science, University of
    Copenhagen\\ \email{robin@di.ku.dk} \and
Department of Computer Science, IT University of
    Copenhagen\\ \email{nive@itu.dk}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
Reversible computation studies computations which exhibit both forward and
backward determinism. Among others, it has been studied for half a century for
its applications in low-power computing, and forms the basis for quantum
computing.

@@

In this paper, we extend the two-level calculus of finitary \Pio{} to one for
full \Pio{} (i.e., with both recursive types and iteration by means of a
\trace{} combinator) using the \Delay{} monad, which can be regarded as a
``computability-aware'' analogue of the usual $\mathsf{Maybe}$ monad for
partiality. This yields a calculus of iterative (and possibly non-terminating)
reversible programs acting on user-defined dynamic data structures together
with a calculus of certified program equivalences between these programs.

@@

\section{TODO} % (fold)
\label{sec:todo}
Robin:
\begin{itemize}
  \item Make pictures! 
  \begin{itemize}
    \item Hasegawa trace from iterator.
  \end{itemize}
  \item Conclusion
  \begin{itemize}
    \item Related work: Kleisli category of the Delay monad is a join
    restriction category, inverse category of a join restriction category is a
    join inverse category; already known to be dagger-traced under certain
    conditions.
    \item Future work: Semantics of time invertible programming. Extension of
    the work on the Kleisli category of the Delay monad beyond $\Set$ to other
    (more general) categories, e.g. topoi.
  \end{itemize}
\end{itemize}
Niccolò: 
\begin{itemize}
  \item Fix the syntax to be closer to ordinary presentations (but do describe 
  how it is realized in Agda).
  \item Finish the section about trace and iteration.
  \item Remind the reader about double lines of coinductive definitions.
  \item Swap types in Elgot iterator.
\end{itemize}
Both:
\begin{itemize}
  \item \textbf{Important:} Come up with a punnier title! (Something about 
  delays and daggers).
\end{itemize}
% section todo (end)

\section{Introduction}\label{sec:intro}
Reversible computation is an emerging computation paradigm encompassing
computations that are not just deterministic when executed the \emph{forward}
direction, but also in the \emph{backward} direction. While this may seem
initially obscure, reversible computation forms the basis for quantum
computing, and has seen applications in a number of different areas such as
low-power computing~\cite{Landauer61}, robotics~\cite{SchultzLE18}, discrete
event simulation~\cite{Schordan15}, and the simultaneous construction of
parser/pretty printer pairs~\cite{RendelO10}. Like classical computing, it has
its own automata~\cite{Bennett73}, circuit model~\cite{DeVos10}, machine
architectures~\cite{ThomsenAG11}, programming
languages~\cite{JamesS12,JamesS14,YokoyamaG07,Schultz18,JacobsenKT18}, semantic
metalanguages~\cite{Giles,KaarsgaardAG17,KaarsgaardG18}, and so on.

@@

Throughout the paper, we reason constructively in Martin-L\"of type
theory. Classically, the delay monad is isomorphic to the maybe monad
$\Maybe\,X \equiv X + 1$,
and thus just a complication of something that can be expressed much
simpler. Constructively, however, they are very different. In particular,
it is impossible to define a well-behaved trace combinator for the
maybe monad without assuming classical principles such as the law of
excluded middle.

We have fully formalized the development of the paper in the
dependently typed programming language Agda~\cite{Agda}. The code is
available online at \\ \texttt{\url{https://github.com/niccoloveltri/pi0-agda}}. The
formalization uses Agda 2.6.0.
